Cycle Route Uptake and Scenario Estimation (CRUSE): an approach for developing strategic cycle network planning tools

### 2024-03-26 paper.qmd
add tbl-trip_purposes

---
title: "Cycle Route Uptake and Scenario Estimation (CRUSE): an approach for developing strategic cycle network planning tools"
# subtitle: A case study of Ireland
bibliography: references.bib
csl: apa.csl
editor:
  markdown:
    wrap: sentence
execute: 
  echo: false
  message: false
number-sections: true
latex-clean: false
pdf-engine: pdflatex
knitr:
  opts_chunk:
    fig.path: images/
format: 
  # html: default
  # docx: default
# # # Uncomment for formatted version
  elsevier-pdf:
    linestretch: 2
    keep-tex: true
    journal:
      name: European Transport Research Review
      formatting: preprint
      model: 3p
      cite-style: super
# # # #   # pdf: default

# # Uncomment the following line to produce the cover page
author:
  - name: Robin Lovelace
    email: r.lovelace@leeds.ac.uk
    corresponding: true
    affiliations:
      - name: University of Leeds
        address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
  - name: Joey Talbot
    affiliations:
      - name: University of Leeds
        address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
  - name: Eugeni Vidal-Tortosa
    affiliations:
      - name: University of Leeds
        address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
  - name: Hussein Mahfouz
    affiliations:
      - name: University of Leeds
        address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
  - name: Elaine Brick
    affiliations:
      - name: AECOM
        address: Unit 6, Galway Technology Park, Parkmore, Galway, Ireland
  - name: Peter Wright
    affiliations:
      - name: AECOM
        address: Winslade House, Winslade Park, Manor Drive, Clyst St Mary, Exeter, EX5 1FY, UK
  - name: "Gary O'Toole"
    affiliations:
      - name: AECOM
        address: Unit 6, Galway Technology Park, Parkmore, Galway, Ireland
  - name: Dan Brennan
    affiliations:
      - name: Transport Infrastructure Ireland
        address: Parkgate Business Centre, Parkgate Street, Dublin 8, D08 DK10, Ireland
  - name: Suzanne Meade
    affiliations:
      - name: Transport Infrastructure Ireland
        address: Parkgate Business Centre, Parkgate Street, Dublin 8, D08 DK10, Ireland
abstract: |
  This paper describes an approach for developing strategic cycle network planning tools. Based on our experience developing and deploying the Cycle Route Uptake and Scenario Estimation (CRUSE) Tool for Ireland, we outline the underlying methods, including disaggregation of origin-destination data, incorporation of additional trip purposes, routing, scenario generation, and development of an intuitive user interface. Commissioned and used by Transport Infrastructure Ireland, CRUSE provides estimates of current and potential future cycling levels under 'snapshot' scenarios to inform investment decisions. The publicly available results at https://cruse.bike/ enable planners, engineers, and other stakeholders to make more evidence-based decisions. CRUSE goes beyond previous work by: modeling networks at high spatial resolution; simulating multiple trip purposes (social, shopping, personal utility, recreational, and cycle touring) in addition to using official origin-destination data; and providing estimates of 'quietness' (a proxy for cyclist comfort and route preference) at the route segment level. Three network types --- 'Fastest', 'Balanced', and 'Quietest' --- help plan both arterial and residential cycle networks. Workshops and unstructured interviews with stakeholders show that the tool is already being used in practice to inform urban, inter-urban, and rural network planning. The approach is flexible and open-source, allowing the underlying ideas and code to be adapted, supporting effective cycling policies and interventions internationally.
keywords: 
  - Cycling
  - Open-source
  - Road Safety
  - Active Travel
  - Transport Planning
  - Collaborative Planning
date: last-modified
---

@@

## Background

Transport systems dominated by heavy and powerful private cars are inefficient, dangerous, and unhealthy.
Cars are a major and growing source of greenhouse gas (GHG) emissions [@winkler2023], a leading cause of premature death and injury due to road traffic collisions [@globals2018], and a cause of disease due to physical inactivity and air, noise and microplastic pollution [@mattsson2023] [@welch2023] [@cavallaro2024].
<!-- Shouldn't these three sources share the brackets? --> Transport is responsible for 23% of GHG emissions, 70% of which are from road transport, with passenger cars accounting for nearly half of transport emissions (around 10% of global emissions) [@jaramillo2022].
The transport system encourages, enables and in some cases enforces unsustainable lifestyles.
Services that are only accessible by car lock-in car dependency [@gray2001; @shergold2012; @motte-baumvol2010].

@@

Building cycle networks represents a relatively 'quick win' within the context of decarbonization [@brand2020] and sustainable mobility [@burns2013].
Although cycling uptake appears on the surface to only relate to the 'shift' part of the ASI framework, closer consideration of the knock-on impacts of cycling uptake shows that it can also help avoid unnecessary trips [@nello-deakin2020].
Furthermore, highly efficient ebikes --- which are seeing rapid uptake --- outperform electric cars, which are too heavy and expensive, for the majority of trips.
Over-reliance on electric cars could slow the transition away from car dependency and inadvertently enable "high travel lock-in" [@anable2019].
At the European level, the European Union has a target of reducing GHG emissions by 55% by 2030, compared to 1990 levels, and to achieve 'net-zero' by 2050 [@rosenow2022].
Climate change mitigation is a major motivation for cycle network plans [@scappini2022].

@@

At the regional level within Ireland, the recently published Greater Dublin Area Transport Strategy [@greater] reveals the high support for and capability for cycling: "nearly a quarter of adults cycle at least once a week in the Dublin Metropolitan Area" with cycling in the Dublin area taking up to 60,000 cars off the road today.
Extrapolating this on a per population basis across Ireland, with around 40% of the population living in Dublin, suggests that around 150,000 cars could be removed nationwide just by achieving Dublin levels of cycling in all counties (notwithstanding existing cycling trips and differences in trip distances).
Seven in ten trips in Ireland are by car [@national].
Cycling has the potential to replace a large proportion of these trips: "A high priority must also be given to cyclists, because trips by this mode have the potential to replace trips by private car, most specifically for short to medium distance trips, but increasingly for longer trips as ebikes extend the range of this mode" [@greater].

@@

## Aim and content

This paper describes an approach to strategic cycle network planning tool development that is open, scalable, and evidence-based.
We label the approach Cycle Route Uptake and Scenario Estimation (CRUSE) and present a case study of its application in Ireland, the results of which are publicly available at [cruse.bike](https://cruse.bike/).

In @sec-tools, we review existing current tools for estimating cycling potential.
In @sec-methods, we outline the methods used to generate the evidence presented in the CRUSE Tool.
In @sec-results, we present the results of the CRUSE Tool, including estimates of current cycling levels and future cycling potential at the national, regional and local levels.
In @sec-discussion, we discuss limitations and possible future improvements to the approach, and the implications of the results for Ireland before making concluding remarks in @sec-conclusions.

# Tools for estimating cycling potential {#sec-tools}

@@

Route-based tools emerged with increasing availability of origin-destination data, routing engines that can assign trips to networks, and improvements in computer hardware and software needed to generate route networks.
A prominent example is the the Propensity to Cycle Tool (PCT), first developed to estimate current and future levels of cycling at desire line, zone, route, and route network levels for case study cities [@lovelace_mapping_2016].
The approach was scaled-up to estimate estimate the potential benefits of uptake at zone and desire line levels nationally, and launched as a publicly available web application in 2017 [@lovelace2017].
Extensions of the PCT approach have included estimation of benefits at the individual level [@woodcock2018], addition of travel to school network [@goodman2019], and improved modeling of impacts on health, environmental and distributional outcomes [@woodcock2021].
Initially developed just for England, the PCT was extended to cover all of Wales (for commuter data only) in 2018.

@@

A feature of active travel interventions is that they require dense networks of routes to be effective [@parkin2018].
This means that data with high levels of geographic resolution are necessary to estimate cyclical potential.
However, datasets on travel patterns are often only available at the level of administrative zones.
The Central Statistics Office (CSO) in Ireland provides Place of Work, School or College Census of Anonymized Records ([POWSCAR](https://www.cso.ie/en/census/census2016reports/powscar/)) data on the number of people traveling to work and school at the Electoral Division (ED) level, for example.

In the PCT, the method used to convert OD data to route networks was to calculate a single route between the population weighted centroids of the zones associated with each OD pair.
This method works fine when the OD data represents movement between small areas, but was not appropriate for generating route networks from the POWSCAR data because zone centroids are so far apart that the resulting route networks would be sparse and unrealistic.
To tackle this issue we developed a new method for OD data disaggregation called 'jittering'.
<!--This reference should be removed to anonymise the paper. They will guess that Lovelace is involved, but saying we developed and then citing is as if we did not remove authors names from the paper :) --> The method works by first disaggregating the OD data based on a 'disaggregation threshold' and then assigning each disaggregated 'sub-OD' pair to 'subpoints' within each zone.
As shown in @fig-dublin, the resulting route networks are dense, even in rural areas.

@@

However, travel to school and work account for less than half (39%) of all trips in Ireland, excluding 'returning home' trips, according to the 2022 National Household Travel Survey (NHTS) [@national2022].
This represents a substantial decline in the proportion of trips covered by POWSCAR data compared with pre-COVID survey results, with 51% of non-return trips being for work or education according to the 2017 release of the same report [@national2017].
To tackle this issue, we developed a spatial interaction modeling methodology to estimate the number of trips between each OD pair for additional trip purposes.
The classification of trip purposes used in the CRUSE Tool was guided mainly by the trip purpose classification found within the National Household Travel Survey (NHTS), but with the addition of categories based on the comprehensive POWSCAR data, and the need to include recreational trips and multi-stage trips.
<!-- This is already implemented, right?--> An overview of the trip purposes used in CRUSE is presented in @fig-trip-purposes.
<!--This should be Table 2--->

::: {#fig-trip-purposes}
![](images/paste-4.png)

Everyday trip purposes used in the CRUSE Tool.
:::

@@

- Domestic with cycle trips starting at home
- Domestic trips involving travel by other mode to the start of the cycle trip
- Cycle tourism, where start and end trips are different
- Sport cycling 
-->
```
Following feedback from stakeholders, we added another, 'non-everyday' trip purpose in an extension phase: cycle tourism.
As outlined in the CRUSE Extension report [reference and link to be added on publication], this involved developing an inter-county spatial interaction model to estimate the number of trips between each county for cycle tourism, with trip attractors including campsites and major international transport hubs.

## Routing

@@

The Climate Action Plan scenario models the transport emissions reductions targeted in the Irish Government's [Climate Action Plan 2021](https://www.gov.ie/en/publication/6223e-climate-action-plan-2021/), which aims for a 51% cut in overall GHG emissions by 2030 as part of the pathway to net-zero emissions by 2050.
For transport, this includes 500,000 extra walking, cycling and public transport trips per day by 2030.
In terms of car travel, the target is to "Increas\[e\] the proportion of kilometers driven by passenger electric cars to between 40 and 45% by 2030, in addition to a reduction of 10% in kilometers driven by the remaining internal combustion engine cars." This equates to a 5.5 to 6% reduction in total car km driven.

To model this decrease in car km driven, cycling uptake increases in line with the Go Dutch scenario.
However, we only model shift from driving to cycling.
There is no shift from other modes of transport to cycling.

@@

Under the Go Dutch scenario cycling reaches levels equivalent to those found in the Netherlands, taking account of the effects of route hilliness (measured as mean gradient) and route distance.
This scenario uses the same model as the PCT, allowing trips to shift from any other mode to cycling [@lovelace2017].

@@

The CRUSE web application is statically hosted, meaning that it can be accessed without the need for a server running software such as the R package 'shiny' or Python packages such as 'streamlit' or 'flask' in the background [@wickham2021].
Its landing page is illustrated in @fig-landing).

Typical intended user stories are illustrated in @fig-user-stories, which shows that the tool is designed to be used by both professional and non-professional users.
For the main target audience, professional transport planners working at the county level, the tool provides a range of outputs, including estimates of cycling potential at the county and network level.
The provision of balanced (the default), quietest and fastest route networks enables planners to consider the trade-offs between directness and 'cycle friendliness' when planning new infrastructure.
The tool also provides data downloads, enabling estimates of cycling potential on the network to be visualised and analysed with other tools such as QGIS, Python or R.

@@

Aside from the widely varying sizes and route networks of each city, with Dublin having by far the densest network, it is clear from the visualisation of cycling friendliness that there are many gaps in the networks.
Dublin has had received most investment in cycling infrastructure and this is apparent from the high proportion of the route network that is green, even in the fastest route network representing routes that prioritize speed and directness over comfort and safety.
Dublin also has the highest mode share of cycling of the four cities, adding to the evidence base showing that investment in cycling infrastructure is effective in increasing cycling uptake.

Although Dublin City has made progress, there are still many parts of the fastest route network, and even some parts of the Balanced and Quietest route networks, that are not cycle friendly and which have high cycling potential.
According to a recent report, 71% of residents in the Dublin Metropolitan Area support "more cycle tracks along roads, physically separated from traffic and pedestrians" [@walking2021].
The results for the Dublin area could help prioritise investment in cycling infrastructure in the city.

<!-- The results for the other -->

## Stakeholder engagement and feedback {#sec-qualitative}

@@

Practitioners in Kildare County Council that the evidence generated *was already valuable* the development of new cycle routes, especially when making the case during planning applications.
They emphasized that cycle networks are often fragmented, and the tool allows them to demonstrate how individual local schemes can contribute to a larger, interconnected network.
An unexpected piece of feedback from the workshop was the preference for visualizations that directly compare the fast and quiet routes, as they recognize the importance of both types of routes.
In response to this we provided the 'Route types' page for each county, although this functionality is not available in the landing page map due to other feedback requesting a simpler interface.
In comparison with another tool (the National Transport Authority's cycling propensity tool), the CRUSE tool was seen by practitioners in Kildare as more user-friendly and intuitive, with the ability to generate evidence at the route level, which is crucial for planning new cycling infrastructure.
Overall, the workshop participants concluded that the approach filled gap in evidence and will greatly assist in the development of new cycling projects.

@@

In a subsequent interview with a practitioner using the evidence generated by the CRUSE approach in their everyday work, we received the following feedback.
The outputs were a "great tool for getting started with project outline documents" and "ideal for generating a strategic rationale for estimating demand and building the business case in the absence of cycle count data" (Declan Keenan, Senior Transport Planner working on active travel, personal communication).
Declan also noted the alignment between local knowledge, cycle count data, and the outputs of the approach: "It's a good representation of what's going on" and the networks allign well in terms of network "shape" and the proportion of trips of trips on different parts of the network.

Findings from the qualitative results also suggest changes which have not been implemented due to time constraints, including the provision of route netework data showing only "gold standard" POWSCAR data (we did make this data available in the data downloads page), the need for further validation of results (especially in rural areas where counter datasets are scarce) and the need for training on the tool to ensure that it is well-understood and used effectively.
These and other potential areas for further research and development are discussed in the next section.

@@

The aim of the paper was to describe the design, features and potential use of the 'CRUSE' approach and resulting tool.
As outlined in the previous section, the CRUSE Tool provides evidence on current and potential future cycling levels across Ireland down to the street level, with potential assigned to fastest, balanced and quietest route networks.
This provision of multiple scenarios of behavior change *and* multiple scenarios of investment, for example in cycling infrastructure next to major roads vs quiet residential streets, is a key feature of the CRUSE Tool.

@@

As the benefits of strategic planning for cycling become more apparent [@scappini2022], we expect the demand for tools such as CRUSE to grow.
However, the approach is not without limitations and does not meet all requirements in Ireland or indeed any country where the methods are applied.

## Limitations and future research

The approach has a number of limitations that should be understood by practitioners, researchers, advocates, policy makers using the tool.
As highlighted in @sec-qualitative, we were only able to compare the results with a small number of cycle counters and based on feedback from workshops.
As outlined in the CRUSE extension report [reference to be added on publication], we also undertook validation against Strava data.
However, more data is needed to better understand the accuracy of the baseline results, especially for the 'non-POWSCAR' purposes.
We hope this becomes possible as Ireland's cycle counter network expands, allowing refinements to the approach (espcially the spatial interaction model and parameters) to be made.
Changes that could help mitigate this limitation on the user interface side include providing results with confidence intervals or as categories of cycling potential, rather than as central estimates, and the visualistion of the 'POWSCAR' network (which is based on Census data) in the web application.

A limitation associated with the approach is its reliance on OpenStreetMap (OSM) data.
The 'cycle friendliness' of a route is based on the OSM tags and the routes themselves are generated by the CycleStreets routing engine, which is based on OSM data.
During a conference in Sligo where we presented prelimary results, a stakeholder with local knowledge pointed out that a quiet route along the canal was not being followed.
In response, manual edits were made to the OSM network, and the results currently on the website reflect these changes.
However, there are many other potential errors in the OSM data that have not been corrected.
Future work could mitigate this by integrating the approach more closely with OSM (for example by providing links to enable one-click edits of OSM when users click on a route segment) or by setting up a workshop to 'crowdsource' corrections to the OSM data.
Furthermore, because there are multiple possible users of the tool with different starging points, workshops and training events could be valuable if not essential to ensure that the tool is used effectively.
Such events could bring together stakeholders from advocacy groups, local authorities, and national governments.
The siloed nature of transport planning means that such diverse stakeholders are rarely the same room, meaning that a 'wide boundary' outcome of the tool could be to bring these groups together (if workshops based on the tool are successful).

Another limitation of the approach is that the results are 'static', reducing its ability to support monitoring of growth in cycling, upgrades of existing cycling infrastructure and dynamic (year-by-year) exposure information to estimate crash rates.
Exposure information is a key part of the the European RISM Directive, for reporting collision rates of vulnerable road users by 2024.
These factors mean that the tool can be seen as an open access 'leverage point', providing key information for and supporting ambitious plans in many aspects the planning system, from network design to post-build monitoring [@lovelace2020].

>>>>

---
title: "Cycle Route Uptake and Scenario Estimation (CRUSE): an approach for developing strategic cycle network planning tools"
# subtitle: A case study of Ireland
bibliography:
  # cruse.bib
  - bibliography.bib
  - cruse.bib
  - references.bib
csl: apa.csl
editor:
  markdown:
    wrap: sentence
execute: 
  echo: false
  message: false
number-sections: true
latex-clean: false
pdf-engine: pdflatex
knitr:
  opts_chunk:
    fig.path: images/
format: 
  # html: default
  # docx: default
# # # Uncomment for formatted version
  elsevier-pdf:
    linestretch: 2
    keep-tex: true
    journal:
      name: European Transport Research Review
      formatting: preprint
      model: 3p
      cite-style: super
# # # #   # pdf: default

# # Uncomment the following line to produce the cover page
author:
  - name: Robin Lovelace
    email: r.lovelace@leeds.ac.uk
    corresponding: true
    affiliations:
      - name: University of Leeds
        address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
  - name: Joey Talbot
    affiliations:
      - name: University of Leeds
        address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
  - name: Eugeni Vidal-Tortosa
    affiliations:
      - name: University of Leeds
        address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
  - name: Hussein Mahfouz
    affiliations:
      - name: University of Leeds
        address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
  - name: Elaine Brick
    affiliations:
      - name: AECOM
        address: Unit 6, Galway Technology Park, Parkmore, Galway, Ireland
  - name: Peter Wright
    affiliations:
      - name: AECOM
        address: Winslade House, Winslade Park, Manor Drive, Clyst St Mary, Exeter, EX5 1FY, UK
  - name: "Gary O'Toole"
    affiliations:
      - name: AECOM
        address: Unit 6, Galway Technology Park, Parkmore, Galway, Ireland
  - name: Dan Brennan
    affiliations:
      - name: Transport Infrastructure Ireland
        address: Parkgate Business Centre, Parkgate Street, Dublin 8, D08 DK10, Ireland
  - name: Suzanne Meade
    affiliations:
      - name: Transport Infrastructure Ireland
        address: Parkgate Business Centre, Parkgate Street, Dublin 8, D08 DK10, Ireland
abstract: |
  This paper describes an approach for developing strategic cycle network planning tools. Based on our experience developing and deploying the Cycle Route Uptake and Scenario Estimation (CRUSE) Tool for Ireland, we outline the underlying methods, including disaggregation of origin-destination data, incorporation of additional trip purposes, routing, scenario generation, and development of an intuitive user interface. Commissioned and used by Transport Infrastructure Ireland, CRUSE provides estimates of current and potential future cycling levels under 'snapshot' scenarios to inform investment decisions. The publicly available results at https://cruse.bike/ enable planners, engineers, and other stakeholders to make more evidence-based decisions. CRUSE goes beyond previous work by: modeling networks at high spatial resolution; simulating multiple trip purposes (social, shopping, personal utility, recreational, and cycle touring) in addition to using official origin-destination data; and providing estimates of 'quietness' (a proxy for cyclist comfort and route preference) at the route segment level. Three network types --- 'Fastest', 'Balanced', and 'Quietest' --- help plan both arterial and residential cycle networks. Workshops and unstructured interviews with stakeholders show that the tool is already being used in practice to inform urban, inter-urban, and rural network planning. The approach is flexible and open-source, allowing the underlying ideas and code to be adapted, supporting effective cycling policies and interventions internationally.
keywords: 
  - Cycling
  - Open-source
  - Road Safety
  - Active Travel
date: last-modified
---

@@

Transport systems dominated by heavy and powerful private cars are inefficient, dangerous, and unhealthy.
Cars are a major and growing source of greenhouse gas emissions causing climate change [@winkler2023], a leading cause of premature death and injury due to road traffic collisions [@globals2018] , and a cause of disease due to physical inactivity and air, noise and (from tyres) microplastic pollution [@mattsson2023] [@welch2023] [@cavallaro2024].
Transport is responsible for 23% of global greenhouse gas (GHG) emissions, 70% of which are from road transport, with passenger cars accounting for nearly half of transport emissions (around 10% of global emissions) [@jaramillo2022].
The transport system encourages, enables and in some cases enforces unsustainable lifestyles.
Services that are only accessible by car lock-in car dependency [@gray2001; @shergold2012; @motte-baumvol2010].

@@

Building cycle networks represents a relatively 'quick win' within the context of decarbonization [@brand2020] and sustainable mobility [@burns2013].
Although cycling uptake appears on the surface to only relate to the 'shift' part of the ASI framework, closer consideration of the knock-on impacts of cycling uptake shows that it can also help avoid unnecessary trips [@nello-deakin2020].
Furthermore, highly efficient ebikes --- which are seeing rapid uptake in Ireland and other countries --- outperform electric cars, which are too heavy and expensive.
Over-reliance on electric cars could slow the transition away from car dependency and inadvertently enable "high travel lock-in" [@anable2019].
At the European level, the European Union has a target of reducing GHG emissions by 55% by 2030, compared to 1990 levels, and to achieve 'net-zero' by 2050 [@rosenow2022].
Climate change mitigation is a major motivation for cycle network plans [@scappini2022].

@@

At the regional level within Ireland, the recently published Greater Dublin Area Transport Strategy [@greater] reveals the high support for and capability for cycling: "nearly a quarter of adults cycle at least once a week in the Dublin Metropolitan Area" with cycling in the Dublin area taking up to 60,000 cars off the road today.
Extrapolating this on a per population basis across Ireland, with around 40% of the population living in Dublin, suggests that around 150,000 cars could be removed nationwide just by achieving Dublin levels of cycling in all counties (notwithstanding existing cycling trips and differences in trip distances).
7 in 10 trips in Ireland are by car [@national] .
Cycling has the potential to replace a large proportion of these trips: "A high priority must also be given to cyclists, because trips by this mode have the potential to replace trips by private car, most specifically for short to medium distance trips, but increasingly for longer trips as ebikes extend the range of this mode" [@greater].

@@

This paper describes an approach to strategic cycle network planning tool development that is open, scalable, and evidence-based.
We label the approach Cycle Route Uptake and Scenario Estimation (CRUSE) and present a case study of its application in Ireland, the results of which are publicly available at [cruse.bike](https://cruse.bike/) in the CRUSE Tool for Ireland, the landing page of which is illustrated in @fig-landing.
In @sec-methods, we outline the methods used to generate the evidence presented in the CRUSE Tool.
In @sec-results, we present the results of the CRUSE Tool, including estimates of current cycling levels and future cycling potential at the national, regional and local levels.
In @sec-discussion, we discuss limitations and possible future improvements to the approach, and the implications of the results for Ireland before making concluding remarks in @sec-conclusions.

# Tools for estimating cycling potential

@@

Route-based tools emerged with increasing availability of origin-destination data, routing engines that can assign trips to networks, and improvements in computer hardware and software needed to generate route networks.
A prominent example is the the Propensity to Cycle Tool (PCT), first developed to estimate current and future levels of cycling at desire line, zone, route, and route network levels for case study cities [@lovelace_mapping_2016].
The approach was scaled-up to estimate estimate the potential benefits of uptake at zone and desire line levels nationally, and launched as a publicly available web application in 2017 @lovelace2017.
Extensions of the PCT approach have included estimation of benefits at the individual level [@woodcock2018], addition of travel to school network [@goodman2019], and improved modeling of impacts on health, environmental and distributional outcomes [@woodcock2021].
Initially developed just for England, the PCT was extended to cover all of Wales (for commuter data only) in 2018.

@@

A feature of active travel interventions is that they require dense networks of routes to be effective [@parkin2018].
This means that high levels of geographic resolution are needed in the data used to estimate cycling potential.
However, datasets on travel patterns are often only available at the level of administrative zones.
The Central Statistics Office (CSO) in Ireland provides Place of Work, School or College Census of Anonymized Records ([POWSCAR](https://www.cso.ie/en/census/census2016reports/powscar/)) data on the number of people traveling to work and school at the Electoral Division (ED) level, for example.

In the PCT, the method used to convert OD data to route networks was to calculate a single route between the population weighted centroids of the zones associated with each OD pair.
This method works fine when the OD data represents movement between small areas, but was not appropriate for generating route networks from the POWSCAR data because zone centroids are so far apart that the resulting route networks would be sparse and unrealistic.
To tackle this issue we developed a new method for OD data disaggregation called 'jittering' [@lovelace2022].
The method works by first disaggregating the OD data based on a 'disaggregation threshold' and then assigning each disaggregated 'sub-OD' pair to 'subpoints' within each zone.
As shown in @fig-dublin, the resulting route networks are dense, even in rural areas.

@@

However, travel to school and work only constitute around only 30% of all trips in Ireland.
To overcome this issue, we developed a spatial interaction modeling methodology to estimate the number of trips between each OD pair for additional trip purposes.
The classification of trip purposes used in the CRUSE Tool was guided mainly by the trip purpose classification found within the National Household Travel Survey (NHTS), but with the addition of categories based on the comprehensive POWSCAR data, and the need to include recreational trips and multi-stage trips (not yet implemented).
An overview of the trip purposes used in CRUSE is presented in @tbl-trip_purposes.

```{r}
#| include: false
trip_purposes = read_csv("trip_purposes.csv") %>% 
  mutate_if(is.character, utf8::utf8_encode)
```

```{r}
#| label: tbl-trip_purposes
# For PDF:
knitr::kable(trip_purposes, caption = "Trip purposes used in the CRUSE Tool.", booktabs = TRUE) |>
  kableExtra::kable_styling(latex_options = "striped", full_width = FALSE) |>
  kableExtra::column_spec(1, width = "13em") |>
  kableExtra::column_spec(2, width = "15em") |>
  kableExtra::column_spec(3, width = "12em") |>
  kableExtra::column_spec(4, width = "5em")
# For HTML:
# knitr::kable(trip_purposes, caption = "Trip purposes used in the CRUSE Tool.")
```

@@

- Domestic with cycle trips starting at home
- Domestic trips involving travel by other mode to the start of the cycle trip
- Cycle tourism, where start and end trips are different
- Sport cycling 
-->
```
## Routing

@@

The Climate Action Plan scenario models the transport emissions reductions targeted in the Irish Government's [Climate Action Plan 2021](https://www.gov.ie/en/publication/6223e-climate-action-plan-2021/), which aims for a 51% cut in overall greenhouse gas emissions by 2030 as part of the pathway to net-zero emissions by 2050.
For transport, this includes 500,000 extra walking, cycling and public transport trips per day by 2030.
In terms of car travel, the target is to "Increas\[e\] the proportion of kilometers driven by passenger electric cars to between 40 and 45% by 2030, in addition to a reduction of 10% in kilometers driven by the remaining internal combustion engine cars." This equates to a 5.5 to 6% reduction in total car km driven.

To model this decrease in car driver km, cycling uptake increases in line with the Go Dutch scenario.
However, we only model shift from driving to cycling.
There is no shift from other modes of transport to cycling.

@@

Under the Go Dutch scenario cycling reaches levels equivalent to those found in the Netherlands, taking account of the effects of route hilliness (measured as mean gradient) and route distance.
This scenario uses the same model that was used PCT, allowing trips to shift from any other mode to cycling [@lovelace2017].

@@

The CRUSE web application is statically hosted, meaning that it can be accessed without the need for a server running software such as the R package 'shiny' or Python packages such as 'streamlit' or 'flask' in the background [@wickham2021].
Typical intended user stories are illustrated in @fig-user-stories, which shows that the tool is designed to be used by both professional and non-professional users.
For the main target audience, professional transport planners working at the county level, the tool provides a range of outputs, including estimates of cycling potential at the county and network level.
The provision of balanced (the default), quietest and fastest route networks enables planners to consider the trade-offs between directness and 'cycle friendliness' when planning new infrastructure.
The tool also provides data downloads, enabling estimates of cycling potential on the network to be visualised and analysed with other tools such as QGIS, Python or R.

@@

Aside from the widely varying sizes and route networks of each city, with Dublin having by far the densest network, it is clear from the visualisation of cycling friendliness that there are many gaps in the networks.
Dublin has had received most investment in cycling infrastructure and this is apparent from the high proportion of the route network that is green, even in the 'Fastest' route network representing routes that prioritize speed and directness over comfort and safety.
Dublin also has the highest mode share of cycling of the four cities, adding to the evidence base showing that investment in cycling infrastructure is effective in increasing cycling uptake.

Although Dublin City has made progress, there are still many parts of the Fastest route network, and even some parts of the Balanced and Quietest route networks, that are not cycle friendly and which have high cycling potential.
According to a recent report, 71% of residents in the Dublin Metropolitan Area support "more cycle tracks along roads, physically separated from traffic and pedestrians" [@walking2021].
The results for the Dublin area could help prioritise investment in cycling infrastructure in the city.

<!-- The results for the other -->
## Selected qualitative results {#sec-qualitative}

@@

Practitioners in Kildare County Council that the evidence generated *was already valuable* the development of new cycle routes, especially when making the case during planning applications. They emphasized that cycle networks are often fragmented, and the tool allows them to demonstrate how individual local schemes can contribute to a larger, interconnected network. Council officers expressed a preference for visualizations that directly compare the fast and quiet routes, as they recognize the importance of both types of routes (users can switch between the on the 'Route types' page for each county although this functionality is not available in the landing page map due to other feedback requesting a simpler interface).

In comparison with another tool (the National Transport Authority's cycling propensity tool), the CRUSE tool was seen by practitioners in Kildare as more user-friendly and intuitive, with the ability to generate evidence at the route level, which is crucial for planning new cycling infrastructure.
Overall, the workshop participants concluded that the approach filled gap in evidence and will greatly assist in the development of new cycling projects.

@@

In a subsequent interview with a practitioner using the evidence generated by the CRUSE approach in their everyday work, we received the following feedback.
The outputs were a "great tool for getting started with project outline documents" and "ideal for generating a strategic rationale for estimating demand and building the business case in the absence of cycle count data" (Declan Keenan, Senior Transport Planner working on active travel, personal communication).
Declan also noted the alignment between local knowledge, cycle count data, and the outputs of the approach: "It's a good representation of what's going on" and the networks allign well in terms of network "shape" and the proportion of trips of trips on different parts of the network.
There were two suggestions from Declan which have not yet been implemented: the provision of data showing only "gold standard" POWSCAR data, and the need for training on the tool to ensure that all staff can use it effectively.
These, and other potential improvements, are discussed in the next section.

@@

The aim of the paper was to describe the design, features and potential use of the 'CRUSE' approach and resulting tool.
As outlined in the previous section, the CRUSE Tool provides evidence on current and potential future cycling levels across Ireland down to the street level, with potential assigned to 'fastest', 'balanced' and 'quietest' route networks.
This provision of multiple scenarios of behavior change *and* multiple scenarios of investment, for example in cycling infrastructure next to major roads vs quiet residential streets, is a key feature of the CRUSE Tool.

@@

As the benefits of strategic planning for cycling become more apparent [@scappini2022] we expect the demand for tools such as CRUSE to grow.
However, the approach is not without limitations and does not meet all requirements in Ireland or indeed any country where the methods are applied.

A limitation of the approach as implemented in the CRUSE tool is that the results are 'static', reducing its ability to support monitoring of growth in cycling, upgrades of existing cycling infrastructure and dynamic (year-by-year) exposure information to estimate crash rates.
Exposure information is a key part of the the European RISM Directive, for reporting collision rates of vulnerable road users by 2024.
These factors mean that the tool can be seen as an open access 'leverage point', providing key information for and supporting ambitious plans in many aspects the planning system, from network design to post-build monitoring [@lovelace2020].
